SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2014-03-15 paper/paper.tex
paper wibbles.

@@

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
  opaque  :: a -> Bool
  subVals :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,\footnote{Rather than indexes, an alternative representation is to use a
    zipper data structure~\cite{zipper}; see Section~\ref{sec:conclusions} for a
    discussion.}
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{opaque} is false when the type of the value is an ``interesting
  type''; informally, this is a type other than a primitive type like \ttp{Int}
  or \ttp{Char}.  See Section~\ref{sec:base} for a full discussion.
\item \ttp{subVals} returns a tree of all non opaque-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

The number of constructors in an algebraic data value primarily determines how
large it is.  In some cases, the size of fields that come from primitive
types, like \ttp{Int} or \ttp{Char}, have less of a role in understanding a
counterexample, so in general, we want to prevent our reduction algorithm from
spending time shrinking values from these types.

Consequently, we distinguish a class of types referred to as \emph{opaque
  types}.  If the reduction algorithm encounters a opaque type, it is ignored.
By default, this includes ``primitive'' types (e.g., \ttp{Int}, \ttp{Word},
\ttp{Float}, \ttp{Bool}, and \ttp{Char}).  We also treat recursive container
types like lists specially: lists of opaque types are considered to be opaque
types themselves.  This prevents the algorithm from analyzing, for example,
strings.  A list of non-opaque-types, however, is shrunk by the algorithm.

By providing custom instances, the user can declare any substructure in a data
type to be a opaque type.  Doing so effectively treats values from that type as
``black boxes'' for SmartCheck, making SmartCheck more efficient if the user
knows that some portion of the structure cannot be shrunk or is irrelevant to
the property.

Conversely, if the user wishes to shrink types that are treated as opaque by
default, she can sometimes do so before handing the counterexample to
SmartCheck, say by using QuickCheck's shrink instances to shrink base types
while leaving the remaining structure for SmartCheck to shrink.  In the future,
depending on user experiences, default instances for types can be removed so
that SmartCheck's reduction algorithm does not ignore values from those types.

@@

XMonad passes all of its QuickCheck tests, but let us see what might happen to a
new contributor if things go awry.  Suppose a developer defines a deletion
function to delete a window, if it exists.  There is a deletion function that
exists in XMonad, which is quite complex, given the amount of state that is managed
by \ttp{StackSet}.  However, one function used in deletion is to filter the
stack of windows associated with each workspace defined:
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (/=w) \}
\end{code}
%
\noindent
Now, suppose the programmer makes a simple typo and instead writes
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (==w) \}
\end{code}
%
\noindent
When testing the property \ttp{prop\_delete}, which says that deleting the
focused window of the current stack removes it from the \ttp{StackSet} \ttp{x}.
%
\begin{code}
prop_delete x =
    case peek x of
        Nothing -> True
        Just i  -> not (member i (delete i x))
\end{code}
%
\noindent
QuickCheck returns the large value shown in the introduction.  That value is a
relatively small counterexample, but even the smallest \ttp{StackSet} values are
somewhat visually overwhelming due to the number of fields within it.  Recall
the value returned by SmartCheck after generalization:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
Let us examine what was generalized.  In our test run, we chose to treat data
maps as opaque, so the forth element of \ttp{StackSet} is not generalized, but
is simply the empty map, so looks uninteresting.  The second and third fields of
\ttp{StackSet} are generalized, but the first one is not.  So there is something
particular about it.  So the culprit is one of the small constants (1 and -1) or
having a \ttp{Just} value rather than a \ttp{Nothing}: it turns out that what
matters is having a \ttp{Just} value, which is the stack field that deletion
works on!

>>>>

\todo{need to test with arbitrary for ints turned off.  (Use newtype)}

@@

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
  opaque  :: a -> Bool
  subVals :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at a breadth-first index in a
  value,\footnote{Rather than indexes, an alternative representation is to use a
    zipper data structure~\cite{zipper} to traverse data.  We have chosen
    explicit indexes to write simple tail-recursive algorithms that can easily
    be transcribed to imperative languages.}
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{opaque} is false when the type of the value is an ``interesting
  type''; informally, this is a type other than a primitive type like \ttp{Int}
  or \ttp{Char}.  See Section~\ref{sec:base} for a full discussion.
\item \ttp{subVals} returns a tree of all non opaque-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

SmartCheck focuses on efficiently shrinking and generalizing large data
structures.  It is not intended as a general replacement for QuickCheck's
\ttp{shrink} method.  Consequently, SmartCheck ignores ``primitive'' types
without value constructors, such as \ttp{Char}, \ttp{Int}, and \ttp{Word16}.
Our experience is that for the kinds of properties with counterexamples that
contain massive data structures, shrinking primitive types does not
significantly help in understanding them.  Furthermore, by ignoring these types
by fiat, shrinking time is dependent only on the size of a data structure as
measured by the number of constructors.

We generalize the idea of ignoring primitive types by introducing the concept of
\emph{opaque types}.  If the reduction algorithm encounters a opaque type, it is
ignored.  Opaque types include the primitive types mentioned above, but in the
user can declare any substructure in a data type to be a opaque type by
providing custom instances, .  Doing so effectively treats values from that type
as ``black boxes'', making SmartCheck more efficient if the user knows that some
portion of the structure cannot be shrunk or is irrelevant to the property.

%% The number of constructors in an algebraic data value primarily determines how
%% large it is.  In some cases, the size of fields that come from primitive
%% types, like \ttp{Int} or \ttp{Char}, have less of a role in understanding a
%% counterexample, so in general, we want to prevent our reduction algorithm from
%% spending time shrinking values from these types.

%% By default, this includes ``primitive'' types (e.g., \ttp{Int}, \ttp{Word},
%% \ttp{Float}, \ttp{Bool}, and \ttp{Char}).  We also treat recursive container
%% types like lists specially: lists of opaque types are considered to be opaque
%% types themselves.  This prevents the algorithm from analyzing, for example,
%% strings.  A list of non-opaque-types, however, is shrunk by the algorithm.

%% Conversely, if the user wishes to shrink types that are treated as opaque by
%% default, she can sometimes do so before handing the counterexample to
%% SmartCheck, say by using QuickCheck's shrink instances to shrink base types
%% while leaving the remaining structure for SmartCheck to shrink.  In the future,
%% depending on user experiences, default instances for types can be removed so
%% that SmartCheck's reduction algorithm does not ignore values from those types.

@@

XMonad passes all of its QuickCheck tests, but let us see what might happen to a
new contributor if things go awry.  Suppose a developer defines a deletion
function to delete a window, if it exists.  There is a deletion function that
exists in XMonad, which is quite complex, given the amount of state that is managed
by \ttp{StackSet}.  However, one function used in deletion is to filter the
stack of windows associated with each workspace defined:
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (/= w) \}
\end{code}
%
\noindent
Now, suppose the programmer makes a simple typo and instead writes
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (== w) \}
\end{code}
%
\noindent
When testing the property \ttp{prop\_delete}, which says that deleting the
focused window of the current stack removes it from the \ttp{StackSet} \ttp{x}.
%
\begin{code}
prop_delete x =
    case peek x of
        Nothing -> True
        Just i  -> not (member i (delete i x))
\end{code}
%
\noindent
QuickCheck returns the large value shown in the introduction.  That value is a
relatively small counterexample, but even the smallest \ttp{StackSet} values are
somewhat visually overwhelming due to the number of fields within it.  Recall
the value returned by SmartCheck after generalization:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
Let us examine what was generalized.  In our test run, we chose to treat data
maps as opaque, so the forth element of \ttp{StackSet} is not generalized, but
is simply the empty map, so looks uninteresting.  The second and third fields of
\ttp{StackSet} are generalized, but the first one is not.  So there is something
particular about it.  So the culprit is one of the small constants (1 and -1) or
having a \ttp{Just} value rather than a \ttp{Nothing}: it turns out that what
matters is having a \ttp{Just} value, which is the stack field that deletion
works on!
